Question
========
Is it reasonable to conclude that the point process describing outbreaks of African swine fever across Latvia is homogeneous?

Answerlist
----------
		
* Yes
* No

Solution
========

We first create a ppp object by adapting the code in Section 4.6.4 of the lecture notes:

```{r, warning=FALSE, eval=FALSE}
Latvia <- read_sf( "Shapefiles/gadm41_LVA_0.shp" )
Latvia <- Latvia %>% 
  st_simplify( dTolerance = 1000 ) %>%
  st_transform( crs=3857 )

SwineFever_transformed <- SwineFever_Latvia %>%
  st_as_sf( coords=c("longitude","latitude"), crs=4326 ) %>%
  st_transform( crs=3857 ) %>% 
  st_coordinates( )

Latvia_ppp <- ppp( SwineFever_transformed[,1], 
                   SwineFever_transformed[,2], 
                   window = as.owin(Latvia) )
```

We can then derive and visualize the quadrat counting estimate:

```{r, out.width='60%', fig.align='center', eval=FALSE}
par( mai=c(0.01,0.01,0.5,0.01) )
Latvia_Quadrants <- quadratcount( Latvia_ppp, nx=7, ny=5 )
plot( Latvia_Quadrants, 
      main="Quadrat counting for swine fever outbreaks across Latvia" )
```

From this plot we conclude that there are large spatial differences in the intensity with almost all cases located in north-eastern Latvia. Consequently, we conclude that the process is non-homogeneous.


Meta-information
================
exname: Q1c
extype: schoice
exsolution: 01