Question
========
Which words (exclusing stop words) are not among the five most common words in the book *Frankenstein*?

Answerlist
----------
* life			
* father
* time
* elizabeth			
* mind			
* day	
* the 
* on 

Solution
========

We first split the lines of text and remove any punctuation:

```{r, eval=FALSE, echo=TRUE}
Frankenstein <- Frankenstein_raw %>% unnest_tokens( word, text )
Frankenstein <- Frankenstein %>% mutate( word = gsub( "_", "", word ) )
```

We then count the number of occurrences for each word and remove any words that are categorized as stop words. Finally, we print the five most common words (excluding stop words):

```{r, eval=FALSE, echo=TRUE}
Frankenstein %>% 
  count( word, sort=TRUE ) %>%
  anti_join( stop_words, by=c("word") ) %>% 
  slice_head( n=5 )
```

We conclude that that five most common words (excluding stop words) are “life”, “father”, “eyes”, “time” and “night”.


Meta-information
================
exname: Q3a
extype: mchoice
exsolution: 00011111
exshuffle: 8
