---
title: "MA22019 Introduction to Data Science - Lecture Notes"
author: "Dr Christian Rohrbeck"
date: "February 2025"
bibliography:
- book.bib
- packages.bib
---

# Overview {.unnumbered}

### Content {.unnumbered}

In practice we often have to deal with large and complex data sets (**"big data"**) - many organisations and companies have collected vast amounts of data over the last decade. Analyzing such data and extracting valuable insights from it lies at the heart of data science.

Although there exists no unique definition, the consensus is that data science is interdisciplinary and requires combining expertise across mathematics, computer science and the area of application. As such, we as mathematicians need to have at least some understanding of the context in which the data were collected / are analyzed in order to extract the important information and to communicate effectively with other disciplines.

In this course we will focus on four key areas of data science:

-   Data wrangling

-   Data visualization

-   Text data analysis

-   Spatial data analysis

Throughout this unit we will usually work with real-world data sets which you can download from the MA22019 Moodle page.


### Aims and Objectives {.unnumbered}

After taking this unit, you should be able to:

-   Demonstrate knowledge of data science techniques used for data wrangling and visualization. 

-   Use R for data visualization and wrangling.

-   Show awareness of the applications of these methods. 

-   Understand some of the mathematical models underlying spatial and text data analysis.

-   Apply these methods in R to analyse large and complex data sets. This includes demonstrating a practical and critical approach to data analysis, including the ability to select a suitable data science method, explaining your reason for choosing it, and correctly interpreting the results.


### Prerequisites {.unnumbered}

You should be familiar with the basics of R introduced in the Probability and Statistics component of MA12003. On Moodle, you can find the "Brief introduction to R" provided by Professor Jennison in case you are unsure about certain aspects.

This course will also use some of the fundamentals introduced in Year 1, including the various probability distributions, and the concepts of mean, variance and correlation.


### Assessment {.unnumbered}

#### Summative Assessment {.unnumbered}

Your mark for MA22019 is based on two individual pieces of coursework:

- **Coursework 1: 40% of unit mark.** Set at the end of Week 3 and due at the end of Week 4. This coursework will focus on data wrangling and data visualization.

- **Coursework 2: 60% of unit mark.** Set at the end of Week 10 and due at the start of Revision Week. This coursework will focus on text and spatial data analysis, but you will also have to demonstrate your ability to use data wrangling and data visualization techniques.

You have to submit your solutions to the coursework via Moodle.


#### Formative Assessment {.unnumbered}

Problem sheets will be set at the beginning of the week (except for Weeks 4, 10 and 11) and include:

**Exercises:** 

  : Focus more on the programming aspects of the course and help you to revise the content covered in that week's lectures. You can submit your solutions to most questions via a Moodle quiz and you will receive direct feedback.

**Tutorial questions:** 

  : To be be attempted in the weekly tutorials. Solutions will be made available on Tuesday evening after the last tutorial has finished.
    
**Homework question:** 

  : Open-ended question, similar in style to the coursework questions. Your answer can be submitted via Moodle to your tutor for feedback. After the submission deadline, I will provide a solution which demonstrates the aspects one may consider to address the question (and achieve a first-class mark), but I won't provide a fully formulated answer - I want you to practice discussing your results in your own words. 


### Organisation {.unnumbered}

#### Lecture Notes {.unnumbered}

Lecture notes are available from the MA22019 Moodle page. The notes are available in two formats (HTML and PDF) with identical content but different layouts.

**Important:** While the lecture notes and problem class notes provide most of the relevant information on the mathematics/statistics and programming aspects of the course, they do not fully explore (1) how to decide which plots are best suited for addressing a research question and (2) how to interpret these plots in the context of the application. These skills are best trained by looking at a range of data applications, and we will do so in the lectures and problem classes. The open-ended question on the problem sheets is designed to support your learning and to provide additional practice on this topic. As such, you are strongly encouraged to actively engage with the lectures and problem sheets.


#### Lectures and Tutorials {.unnumbered}

**In person lecture**: Wednesday 10:15-12:05 (Weeks 1-3,5-9) or 10:15-11:05 (Weeks 4,10,11) in 3WN 2.1

**Live online learning session (LOIL)**: Friday 14:15-15:05 (Weeks 1-3) on Zoom (link on Moodle)

**Computer Lab:** You will be assigned to a small group which will meet weekly to go over the tutorial questions on the problem sheet. Please check your timetable to identify the time and location of your tutorial.

### Schedule {.unnumbered}

The rough outline for the individual weeks is as follows:

| Week | Topic                               | Remarks                 |
|:-----|:------------------------------------|:------------------------|
| 1    | Data Wrangling                      | Tutorial on RMarkdown   |
| 2    | Data Wrangling + Data Visualization | Problem Sheet 1         |
| 3    | Data Visualization                  | Problem Sheet 2         |
|      |                                     | Release of Coursework 1 |
| 4    | Q&A Coursework 1                    | Deadline Coursework 1   |
| 5    | Text Data Analysis 1                | No tutorials            |
| 6    | Text Data Analysis 2                | Problem Sheet 3         |
| 7    | Spatial Data Analysis 1             | Problem Sheet 4         |
| 8    | Spatial Data Analysis 2             | Problem Sheet 5         |
| 9    | Spatial Data Analysis 3             | Problem Sheet 6         |
| 10   | Revision Class                      | Release of Coursework 2 |
| 11   | Q&A Coursework 2                    |                         |
| 12   |                                     | Deadline Coursework 2   |  

We will finish the core content by the end of Week 9. In Weeks 4, 10 and 11, we will only have Revision and Coursework Q&A sessions.

### R and RStudio {.unnumbered}

We will use R and RStudio throughout this course. You can access RStudio via the <a href="https://rdweb.wvd.microsoft.com/arm/webclient/index.html">UniDesk</a>, but I would recommend that you install R and RStudio on your own computer / laptop. If you already have R installed, please check that you have at least version 4.4.0.

To install all the relevant packages, run the R code in the file "InstallPackages.R", which you can find on Moodle. The individual packages are then loaded using the **library()** function, e.g.

```{r, eval=F, echo=T}
library(dplyr) 
```

This has to be done every time you start R/RStudio.

To load data from external files, we will use the following two functions

```{r, eval=FALSE, echo=TRUE}
load( "..." )     # To load .RData files
read.csv( "..." ) # To load .csv files
```



### Resources {.unnumbered}

This unit is self-contained in the sense that you will not need to read text books. However, you may wish to consult the following books, available as ebooks via the University Library, to support your learning and understanding:

Boehmke, Bradley C. *Data Wrangling with R.* 1st Edition. 2016. Springer.

Mailund, Thomas. *Beginning Data Science in R : Data Analysis, Visualization, and Modelling for the Data Scientist.* 1st Edition 2017. Apress.

Wickham, Hadley. *ggplot2 : Elegant Graphics for Data Analysis.* 2nd Edition. 2016. Springer.

Silge, Julia and Robinson, David. *Text Mining with R: A Tidy Approach.* 2015. O'Reilly.

**Remark:** Some examples in these books use R functions that have been superseded, i.e., it is advised to use an alternative, more recent function (which may have a different syntax). I will use the most recent functions in this course.
