---
title: "Homework 3"
subtitle: "Mise en place"
description: "Due on ..."
categories: "Homework"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### **Overview**


Your answer to the Homework Question can be submitted on Moodle to your tutor for feedback. The submission deadline is 17:00 on Thursday 20 February 2025. You should submit a single Word, PDF or HTML file that provides your R code, any created R output and all your comments.

Before starting the questions, please make sure to load the relevant R packages: 

```{r, warning=FALSE, message=FALSE}
library(dplyr)
library(lubridate)
library(ggplot2)
library(patchwork)
```


### **Homework Question - Optimizing growing conditions for orchids**

A research institute set up an experiment in 2024 to determine the best growing conditions for two types of Orchids: Cymbidium and Dendrobium. The design of the experiment was as follows:

* All orchids were planted in March or April 2024. Each individual orchid was exposed to constant temperature and phosphate levels, but levels differed across orchids. 

* The height (in inches) and quality of each orchid were measured on 20 October 2024. Plant quality was assessed using a score between 1 and 10 (1=very poor, 10=excellent). Any orchid with a score above 6 is considered of "good" quality.

The research institute approached us to analyze their collected data which provides:

* **Type** - Type of the orchid.

* **Height** - Height of the orchid on 22 October 2024.

* **Quality** - Quality as measured on 22 October 2024.

* **Phosphate** - The level of phosphate in ppm (parts per million) the orchid was exposed to since it had been planted.

* **Temperature** - The temperature (in degree Celsius) the orchid was exposed to since it had been planted.

* **Planting** - The date the orchid was planted.

The full data are provided in the file “Orchids.csv” and the research institute is interested in:

a) How do the two types of orchids compare in terms of the heights measured on 22 October 2024?

b) What are the effects of phosphate and temperature on the height of the orchids?

c) Did the date when the orchid was planted have any effect on whether their quality was at least "good" on 22 October 2024?

Perform an analysis which considers the three aspects above. Make sure to clearly state your approach and conclusions.