---
title: "MA22019 2025 - Computer Lab 9"
subtitle: "Mise en place"
description: "To work on the lab on date xx"
categories: "Lab"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### **Overview**

 Tutorial Questions 1 and 2 provide some more advanced questions. You can check your answers for Exercises 1 and 2 using the quiz provided on Moodle.


You may want to load the following packages before starting the exercise:

```{r, message=FALSE, warning=FALSE}
library( dplyr )
library( ggplot2 )
library( tidyr )
library( sf )
library( ggspatial )
library( prettymapr )
library( spatstat )
```

When working on a University PC, you will have to first install some of these packages. 



### **Tutorial Question 1 - Crime across Utopia**

Utopia's police department needs your help with analyzing their 2015-2021 data regarding certain crimes. The data is provided in the file "UtopiaCrimes.csv". Note, longitude and latitude coordinates are only available for drug possession offences across District 44.

Utopia consists of 59 districts and a shapefile of Utopia is provided as "UtopiaShapefile.shp". To hide Utopia’s location, constants have been added to the latitude and longitude coordinates, but the shapes they define are correct. The population for each district is provided in the file "UtopiaPopulation.csv".

a) Identify the three most common crimes in Utopia for the period 2015-2021. 

b) Visualize the rate per 1,000 population for the most common crime for 2015-2021 for the different districts. What do you conclude from your plot?

c) You are told that District 44 is notorious for drug possession. The police is planning to conduct a raid to tackle the issue, but they are unsure which areas of the district are the most seriously affected. Use spatial data analysis techniques to identify the parts of District 44 they should be focusing on. **Hint:** You may want to consider the function **st_reverse()** should you get an error message regarding orientation/direction.


### **Tutorial Question 2 - Cost of living crisis in Texas**

As many other countries around the world, the United States have seen high inflation over the past years. In the following we want to analyze socioeconomic data for the state of Texas. We have access to two data files:

* "Cost_Texas.csv" gives the cost for essentials (food, rent, healthcare,etc.) in 2023 for a single household for each county in Texas based on the Family Budget Calculator by the Economic Policy Institute.

* "Income_Texas.csv" gives the median income per capita for each county in Texas as reported by the Bureau of Economic Analysis of the US government for 2021.

Perform the following tasks and answer the research questions:

a) Load the two data files and calculate the difference between median income and cost of living for each county.

b) The file "Texas.Rdata" provides a shapefile of Texas, called **Texas**,  with all county boundaries. Use the shapefile to create a map which visualizes the variable calculated in part a). What do you conclude?

c) How reliable are our results found in part b)? **Hint:** You may want to extract the two counties with the highest median income and see what you find out about them on Wikipedia.

