Question
========
What is the difference in average traffic volume between work days and the weekend/ a holiday?

Solution
========
We first perform the data operations from part a):

```{r, eval=FALSE, echo=TRUE}
Traffic <- read.csv( "Traffic Minnesota.csv" )
Traffic <- Traffic %>% mutate( date_time = dmy_hm(date_time) )
Traffic <- Traffic %>% mutate( Weekday = weekdays(date_time) )
```

We then create a variable which represents whether an observation comes from Monday-Friday or whether it's the weekend / a holiday.

```{r, eval=FALSE, echo=TRUE}
Traffic <- Traffic %>% 
  mutate( WorkDay = case_when( Weekday == "Saturday" ~ "Weekend",
                               Weekday == "Sunday" ~ "Weekend",
                               holiday != "None" ~ "Weekend",
                               .default = "Workday" ) )
```

Next, we calculate the average traffic volume for the two categories:
  
```{r, eval=FALSE, echo=TRUE}
AverageTraffic <- Traffic %>% 
  group_by( WorkDay ) %>%
  summarise( Mean = mean(traffic_volume) )
```

Finally, we calculate the difference in the averages:

```{r, eval=FALSE, echo=TRUE}
AverageTraffic$Mean[1] - AverageTraffic$Mean[2]
```

So the difference is 975.372.


Meta-information
================
exname: Q2bi
extype: num
exsolution: 975.372
extol: 0.01
