---
title: "Problem Class 5"
date: "19/03/2025"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background

Sea surface temperature is the temperature of the top millimeter of the ocean's surface. Sea surface temperatures influence weather, including hurricanes, as well as plant and animal life in the ocean. Like Earth's land surface, sea surface temperatures are warmer near the equator and colder near the poles. Currents like giant rivers move warm and cold water around the world's oceans. Some of these currents flow on the surface, and they are obvious in sea surface temperature images.

In this short exercise we consider the average sea surface temperature for the North Atlantic in August 2022 as recorded by <a href="https://neo.gsfc.nasa.gov/view.php?datasetId=MYD28M&date=2022-08-01">NASA</a>. The data is produced by placing a fine grid over the globe and then deriving the average temperature for each pixel.  

Let's import the data

```{r}
library(dplyr)
Sea_raw <- read.csv( "Data/SeaSurface.csv" )
glimpse(Sea_raw)
```

Our aim is to visualize the spatial distribution, and we load the necessary packages

```{r, warning=FALSE, message=FALSE}
library(ggplot2)
library(sf)
```

Consider the following tasks.

## Tasks 

**Task 1:** What is the type of spatial data we are working with?

**Task 2:** Convert the data into a spatial object using the **st_as_sf()** function:

```{r}
Sea <- st_as_sf( Sea_raw, coords=c("lon", "lat"), crs="WGS84" )
```


**Task 3:** Visualize the data using the WGS84 coordinate system. What do you conclude?

```{r}
ggplot( Sea ) + geom_sf( aes(color=Temperature) ) + theme_bw()
```



**Task 4:** Visualize the data using coordinate system with CRS=3347. Which aspects are better represented by this projection?

```{r}
ggplot( Sea ) + geom_sf( aes(color=Temperature) ) + theme_bw() + 
  coord_sf( crs=st_crs( 3347) )
```

