---
title: "MA22019 - Combining data frames"
author: "Christian Rohrbeck"
date: "12 February 2025"
output: 
  ioslides_presentation:
    fig_height: 5
    logo: uob.png
    css: styles.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Plan for Today

So far we have analyzed a single complex data frame using the dplyr R package.

Today we explore 

- how we can combine multiple data frames (Section 1.3)

- the elements of a plot (Sections 2.1.1 and 2.1.2)

- visualizing data using ggplot2 (Section 2.2)


# Combining Data Sets

## Motivation

Let's revisit the river flow data for Bathford

```{r, warning=FALSE, message=FALSE, echo=TRUE}
library(dplyr)
library(lubridate)
Bathford <- read.csv( "Bathford River Flow.csv", skip=20, header=FALSE,
                      colClasses = c("character","numeric","NULL") )  %>%
  rename( Date = V1, RiverFlow = V2 ) %>%
  mutate( Date = as_date( Date, format="%Y-%m-%d" ) )
glimpse( Bathford )
```

## Reporting Missing Data

When we work with real-world data, we have to check whether data are missing.

If there is missing data, you should report the amount of missing data!

```{r, echo=TRUE}
Bathford %>% 
  summarize( "Proportion Missing" = mean( is.na(RiverFlow) ) )
```

We see that river flow is missing for about 0.6% of days.



## Combining data sets with dplyr

**How can we add data from other gauges to the data frame?**

The dplyr package offers four functions to combine data frames based on matching values, which we call the "key(s)".

- inner_join()

- full_join()

- left_join()

- right_join()

The functions differ in terms of observations kept.


## Example - Loading the data

Let's load the river flow data for another gauge:
```{r, echo=TRUE, warning=FALSE}
Bath <- read.csv( "Bath River Flow.csv", skip=19, header=FALSE,
                  colClasses = c("character","numeric","NULL") )
Bath <- Bath %>% rename( Date = V1, RiverFlow = V2 ) %>%
  mutate( Date = as_date( Date, format="%Y-%m-%d" ) )
glimpse( Bath )
```


## Example - Comparing the data sets

Let's check the first elements of each data set

```{r, echo=TRUE}
slice_head( Bath, n=2 )
slice_head( Bathford, n=2 )
```

**What do we notice?**


## Example - Merging the data sets {.build}

<div>
Let's merge observations based on **Date**

```{r, echo=TRUE}
RF <- full_join( Bathford, Bath, by=c("Date" = "Date") )
slice_head(RF, n=1)
```
</div>

<div>
We see that the variable names need updating:

```{r, echo=TRUE}
RF <- rename( RF, Bathford = RiverFlow.x, Bath = RiverFlow.y )
```
</div>

## Example - Data analysis

We explore dependence of the river flow levels

```{r, fig.align='center', out.width='80%', echo=FALSE}
plot( RF$Bath, RF$Bathford, pch=19,
      xlab="River Flow at Bath", ylab="River Flow at Bathford" )
```


## Comparison with inner_join()

While full_join() keeps all observation, inner_join() only keeps the dates listed in both data sets

```{r, echo=TRUE}
RF <- inner_join( Bathford, Bath, by=c("Date" = "Date") )
slice_head(RF, n=3)
```

## Remarks {.build}

<div>
- left_join() and right_join()
</div>

<div>
- Combining a large number of data sets <br> -> **for loop** (Section 1.3.2)
</div>

<div>
- Order of elements after merging

```{r, echo=TRUE}
RF <- full_join( Bath, Bathford, by=c("Date" = "Date") )
slice_head( RF, n=1 )
```
</div>


## Key Messages

We have now finished the chapter on **Data Wrangling**. The take-home messages are

* Ensure that variables are of the correct type and use informative variable names

* Use the dplyr package to structure and manipulate your data

* The restructuring process should be determined by the research question

* Check for missing data and report the proportion if data are missing


