---
title: "MA22019 2025 - Quiz 2"
subtitle: "Mise en place"
description: "To work after the lecture XXX?"
categories: "Quiz"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### **Overview**

Exercises 1-3 will help you with revising the techniques covered in the lectures in Week 1 (Sections 1.1 and 1.2 of the lecture notes). You can check your solutions to these questions yourself by entering them on Moodle via the Quiz provided in the "Problem Sheets" section.


Before starting the exercises, make sure to load the dplyr and lubridate R packages: 

```{r, warning=FALSE, message=FALSE}
library(dplyr)
library(lubridate)
```

### **Exercise 1 - Functions for data wrangling**

We described that data may not be in the format required for the analysis because

-   Variable names are uninformative;

-   Data types are incompatible with available R functions;

-   We are only interested in a subset of the data.

For each of the following R functions, decide which issue they may address and submit your answers via the Moodle quiz:

a) **as_date()**

b) **as.numeric()**

c) **filter()**

d) **group_by()**

e) **rename()**

f) **slice_head()**

If you are unsure about a function, you can use the **help()** function in R, or look at the cheat sheets provided on Moodle.

### **Exercise 2 - Flights from New York City airports**

The file "NYCFlights.csv" provides data for flights in 2013 that departed from any of the three main airports servicing New York City: John F Kennedy (JFK), LaGuardia (LGA) and Newark (EWR). To load the data into your R Workspace, you can use

```{r}
NYCFlights <- read.csv( "data/NYCFlights.csv" )
```

Answer the following questions using the provided data and submit your answers via the Moodle quiz:

a) How many flights does the data set contain?

b) For which month were the most flights recorded?

c) Which plane (specified by the **tailnum** variable) departed most often from New York City airports in 2013?

d) How often did the plane identified in part c) depart from New York City airports in January?


### **Exercise 3 - Earthquakes around the world**

The data set "Earthquakes.csv" contains the location and size of all significant earthquakes, as recorded by the National Earthquake Information Center (NEIC), which occurred worldwide between 1965 and 2016.

We can load the data using 

```{r}
Earthquakes <- read.csv( "data/Earthquakes.csv" )
```

Answer the following questions and submit your answers via the Moodle quiz:

a) In which year between 1965 and 2016 were the most earthquakes recorded?

b) Were more earthquakes recorded in the northern or the southern hemisphere between 1965 and 2016?


