Question
========
Which of the following statements is correct when you explore the proportions with which the different topics feature?

Answerlist
----------
* Topic 1 is featuring significantly more in articles in the "business" category
* Topic 1 is featuring significantly less in articles in the "business" category
* There is no real difference between the "sports" and  "business" category in terms of how often Topic 1 features


Solution
========

We first extract the estimated proportions:

```{r, eval=FALSE}
Articles_topics <- tidy( Articles_LDA , matrix = "gamma" ) %>% 
  separate( document, c("NewsType", "ID"), sep="_", convert=TRUE )
```

We can now create a box plot of proportions for each of the articles

```{r,eval=FALSE}
ggplot( Articles_topics, aes( x=factor(topic), y=gamma ) ) + 
  facet_wrap( ~NewsType ) + geom_boxplot() + 
  labs( x="Topic", y="Proportion" )
```


We find that articles from "business" category are usually made up of 25% Topic 1 and 75% Topic 2. The reverse is true for the "sports" category. So we find that that Topic 1 is featuring significantly less in articles in the "business" category.

Meta-information
================
exname: Q2b
extype: schoice
exsolution: 010
