---
title: "MA22019 2025 - Homework 7"
subtitle: "Mise en place"
description: "Due on ..."
categories: "Homework"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### **Overview**

The Homework question is inspired by one of last year's coursework questions. Besides the creation of plots, the question asks you to carefully discuss the assumptions we make when applying the tools covered in Sections 4.1 and 4.2.

Your answer to the Homework Question can be submitted on Moodle to your tutor for feedback. The submission deadline is 17:00 on Thursday 27 March 2025. You should submit a single Word, PDF or HTML file that provides your R code, any created R output and all your comments.

You may want to load the following packages before starting the exercises:

```{r, message=FALSE, warning=FALSE}
library( dplyr )
library( ggplot2 )
library( sf )
library( ggspatial )
library( prettymapr )
```



## **Homework Question - Lead concentration across Amaurot**

The local authorities in Amaurot, the capital of Utopia, have seen an alarming increase in lead concentration levels in the drinking water. They thus collected samples on lead concentration across the city. The data on the recorded lead levels and the spatial coordinates are provided in the file "Amaurot Lead.csv".

The local authorities have now approached you to help them tackle the issue. They provided you with the collected data and a shapefile of Amaurot. To hide Amaurot’s location, the latitude and longitude coordinates have been manipulated, but the provided shapes are correct. You are asked to perform the following tasks:

a) Visualize the measured lead concentrations. What do you conclude?

b) Perform inverse distance weighting to predict lead levels for all points in the file "Amaurot Grid.csv"; the points form a regular grid over the whole city and there is also information in which district the point lies. Comment on the reliability of your predictions.

c) The authorities want to reduce the occurrences of lead concentrations exceeding a level of 10 parts per million. They have thus decided to deploy a team to one of the districts with the job of reducing lead levels to below this threshold for all households in that districts. Use your results in part b) to identify which district should be targeted.