---
title: "MA22019 2025 - Homework 5"
subtitle: "Mise en place"
description: "Due on "
categories: "Homework"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### **Overview**


Your answer to the Homework Question can be submitted on Moodle to your tutor for feedback. The submission deadline is 17:00 on Thursday 13 March 2025. You should submit a single PDF or Word file that provides your R code, any created R output and all your comments.

You may want to load the following packages before starting the exercise:

```{r, message=FALSE, warning=FALSE}
library( dplyr )
library( ggplot2 )
library( tidytext )
library( wordcloud )
library( stringr )
```

When working on a University PC, you have to first install the tidytext and wordcloud packages. The installation of the textdata package is quite cumbersome, and thus I provide the AFINN and Bing sentiment lexicons as .csv files for your convenience. You can load them using

```{r}
AFINN <- read.csv("data/AFINN Sentiment Lexicon.csv")
Bing <- read.csv("data/Bing Sentiment Lexicon.csv")
```




### **Homework Question - The work of H.G. Wells**

We want to analyze and compare the novels *The Time Machine* and *War of the Worlds* by the British author Herbert George Wells. The books are provided in the files "Time Machine.csv" and "War of Worlds.csv".

a) Compare the two books in terms of the words used within them.

b) How do "The Time Machine" and "The War of the Worlds" differ regarding their sentiment?
