---
title: "Homework 1: R Basics"
author: "Your Name Here"
date: "2026-02-15"
format: html
---

# Instructions

*   **Due Date**: Friday Week 3 (Feb 15) at 17:00.
*   **Setup**: Run `source("check_location.R")` in the Console to verify your project location.
*   **Submission**: Commit and Push your work to GitHub.
*   **Grading**: This is a *formative* assignment (feedback only, no marks).
*   **Note**: The `_files/` folder generated when you render contains temporary files. It is intentionally ignored by Git to keep your repository clean. You do not need to push it.

---

## Question 1: Data Manipulation

Load the `mtcars` dataset and calculate the average horsepower (`hp`) for cars with 4, 6, and 8 cylinders (`cyl`).

```{r}
# Your code here
library(dplyr)
data(mtcars)

# Hint: Use group_by() and summarize()
```

## Question 2: Visualization

Create a scatter plot of Weight (`wt`) vs Miles Per Gallon (`mpg`). Color the points by the number of Gears (`gear`).

```{r}
# Your code here
library(ggplot2)

```

## Question 3: Reflection

Did you find any part of this assignment difficult? (Type your answer below)

> Write your answer here...
