---
title: "Homework 2"
subtitle: "Mise en place"
description: "Due: End of lab on Mon, Aug 25"
categories: "Homework"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### **Overview**


Your answer to the Homework Question can be submitted on Moodle to your tutor for feedback. The submission deadline is 17:00 on Thursday 13 February 2025. You should submit a single Word, PDF or HTML file that provides your R code, any created R output and all your comments.

Before starting the exercises, make sure to load the dplyr and lubridate R packages: 

```{r, warning=FALSE, message=FALSE}
library(dplyr)
library(lubridate)
```

If you are using your own laptop, you may have to first run the code in "InstallPackages.R" (available from Moodle) to install all the packages essential for this course.


### **Homework Question - Salaries for data scientists**

The file "Salary Data Scientists.csv" provides data for 603 employees working in Data Science / Machine Learning / AI. For each employee, we are given the following information

* **work_year** - Year the data were collected for the employee

* **experience_level** - Experience Level: Entry-level (EN), Junior (MI), Senior (SE) or Expert (EX)

* **employment_type** - Full-time (FT), Part-time (PT) or Contingent Workers (CT)

* **salary, salary_currency** - Nominal salary and currency in which it was paid

* **salary_in_usd** - Nominal salary in U.S. Dollars

* **employee_residence** - Country the employee is residing

* **remote_ratio** - Proportion of time the employee may work from home

* **company_location** - Location of the company's headquarters

* **company_size** - Size of the company: Small (S), Medium (M) or Large (L)

An undergraduate is considering to take units in Data Science / Machine Learning / AI. However, they are unsure about the working conditions they are likely to experience if they decide to work in the sector. Specifically, the student plans to work full-time for an U.S. company, and they want to get some insight into the differences between small-/medium-sized and large companies. After a discussion with the student, the following three aspects have been identified as important for them: 

a) Opportunity to work remotely for at least 50% of the time.

b) Salaries for Entry-Level positions.

c) Structure of the workforce in terms of experience level.

Perform an analysis which considers the three aspects above. Make sure to clearly describe your approach and conclusions.
