Question
========
Which month had the highest median daily maximum temperature according to your plot in part d)?

Answerlist
----------
* January
* February
* March
* April
* May
* June
* July
* August
* September
* October
* November
* December

Solution
========
The first step is to create a new variable in the data frame which represents the month of the observation. We apply the functions mentioned in the question text as follows:

```{r, eval=FALSE, echo=TRUE}
NYC <- NYC %>% mutate( Date = as_date( Date, format="%d/%m/%Y") )
NYC <- NYC %>% mutate( Month = factor( month(Date) ) )
```

We then use the function geom_boxplot() from the lecture to produce for each momth a boxplot for the daily maximum temperature:

```{r, fig.align='center', out.width='60%', fig.height=4, fig.width=6}
ggplot( NYC, aes(x=Month,y=Tmax) ) + geom_boxplot() + 
  labs( x="Month", y="Maximum daily temperature in Fahrenheit" )
```  

Since the box plot visualizes, amongst others, the median value, we conclude that July recorded the highest median in terms of daily maximum temperature.

Meta-information
================
exname: Q3ei
extype: schoice
exsolution: 000000100000
